`r glue::glue('inferCNV results for {params$library_id} with {params$reference_name} reference')`

### 2025-05-07 analyses/infercnv-consensus-cell-type/template-notebooks/SCPCP000015_explore-infercnv-results.Rmd
density plot updates and other miscellany

# immune cell URL
immune_url <- "https://raw.githubusercontent.com/AlexsLemonade/OpenScPCA-analysis/refs/heads/main/analyses/cell-type-consensus/references/consensus-immune-cell-types.tsv"

@@

immune_labels <- readr::read_tsv(immune_url) |>
  dplyr::pull(consensus_annotation)

endo_labels <- c(
  "endothelial cell",
  "blood vessel endothelial cell",
  "microvascular endothelial cell"
)

@@

```{r, fig.height = 10}
# Filter to just cells in the library to avoid warnings
library_has_cnv_df <- has_cnv_df |>
  dplyr::filter(library_id == params$library_id)

celltype_umap <- ggplot(library_has_cnv_df) +
  aes(x = UMAP1, y = UMAP2, color = celltype_n) +
  geom_point(size = 0.1, alpha = 0.3) +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  labs(title = "Cell types") +
  umap_theme

cnv_umap <- ggplot(library_has_cnv_df) +
  aes(x = UMAP1, y = UMAP2, color = total_cnv_per_cell) +
  geom_point(size = 0.1, alpha = 0.3) +
  scale_color_viridis_c() +
  labs(title = "Total CNV per cell") +
  umap_theme

@@

```{r}
cluster_count_df <- has_cnv_df |>
  dplyr::count(subcluster, cell_group_n)
ggplot(cluster_count_df) +
  aes(x = n, fill = cell_group_n) +
  geom_histogram(color = "grey20", bins = 15) +
  facet_wrap(vars(cell_group_n), scales = "free") +
  labs(
    x = "Number of cells in cluster",
    y = "Number of clusters"
  )
```

@@

```{r}
cluster_size_df <- has_cnv_df |>
  dplyr::add_count(subcluster, name = "cluster_size") |>
  dplyr::select(
    total_cnv_per_cell,
    cluster_size
  ) |>
  dplyr::distinct()

@@

ggplot(reference_cells_df) +
  aes(x = subcluster, fill = library_id) +
  geom_bar() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_blank()
  )
```

@@

```{r}
# First, construct data frame with proportion_scaled_ instead of has_cnv
prop_cnv_df <- infercnv_df |>
  dplyr::select(
    library_id,
    barcodes,
    starts_with("proportion_scaled_")
  ) |>
  tidyr::pivot_longer(
    starts_with("proportion_scaled_"),
    names_to = "cnv_type_raw",
    values_to = "proportion"
  ) |>
  # more tidying
  dplyr::mutate(cnv_type_raw = stringr::str_remove(cnv_type_raw, "^proportion_scaled_")) |>
  tidyr::separate(cnv_type_raw, sep = "_", into = c("cnv_type", "chr")) |>
  # remove extra chromosomes and cells not in this library
  dplyr::filter(
    !stringr::str_starts(chr, "chrMT"),
    !stringr::str_starts(chr, "chrGL"),
    library_id == params$library_id
  ) |>
  dplyr::mutate(
    chr = stringr::str_remove(chr, "^chr"),
    # set factor levels
    chr = factor(chr, levels = 1:22),
    chr = forcats::fct_relevel(chr, as.character(validation_df$chromosome))
  )
```

We'll make density plots of the predicted CNV events across chromosomes:
The panels will be ordered first by the chromosomes with known CNVs, and then in order for the rest of the chromosomes.

```{r fig.height=12, fig.width=14}
ggplot(prop_cnv_df) +
  aes(x = proportion, color = cnv_type) +
  geom_density() +
  facet_wrap(
    vars(chr),
    scales = "free",
    ncol = 6
  ) +
  theme(strip.text = element_text(size = 8))
```

>>>>

@@

@@

```{r, fig.height = 8}
# Filter to just cells in the library to avoid warnings
library_has_cnv_df <- has_cnv_df |>
  dplyr::filter(library_id == params$library_id)

celltype_umap <- ggplot(library_has_cnv_df) +
  aes(x = UMAP1, y = UMAP2, color = celltype_n) +
  geom_point(size = 0.25, alpha = 0.4) +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  labs(title = "Cell types") +
  umap_theme

cnv_umap <- ggplot(library_has_cnv_df) +
  aes(x = UMAP1, y = UMAP2, color = total_cnv_per_cell) +
  geom_point(size = 0.25, alpha = 0.4) +
  scale_color_viridis_c() +
  labs(title = "Total CNV per cell") +
  umap_theme

@@

```{r, fig.width = 8}
cluster_count_df <- has_cnv_df |>
  dplyr::count(subcluster, cell_group_n)
ggplot(cluster_count_df) +
  aes(x = n, fill = cell_group_n) +
  geom_histogram(color = "grey20", bins = 15) +
  facet_wrap(vars(cell_group_n), scales = "free") +
  labs(
    x = "Number of cells in cluster",
    y = "Number of clusters"
  )
```

@@

```{r, fig.width = 6}
cluster_size_df <- has_cnv_df |>
  dplyr::add_count(subcluster, name = "cluster_size") |>
  dplyr::select(
    total_cnv_per_cell,
    cluster_size
  ) |>
  dplyr::distinct()

@@

ggplot(reference_cells_df) +
  aes(x = subcluster, fill = library_id) +
  geom_bar() +
  scale_fill_brewer(palette = "Dark2") +
  theme(
    legend.position = "bottom",
    axis.text.x = element_blank()
  )
```

@@

```{r}
# First, construct data frame with proportion_scaled_ instead of has_cnv
prop_cnv_df <- infercnv_df |>
  dplyr::select(
    library_id,
    barcodes,
    starts_with("proportion_scaled_")
  ) |>
  tidyr::pivot_longer(
    starts_with("proportion_scaled_"),
    names_to = "cnv_type_raw",
    values_to = "proportion"
  ) |>
  # more tidying
  dplyr::mutate(cnv_type_raw = stringr::str_remove(cnv_type_raw, "^proportion_scaled_")) |>
  tidyr::separate(cnv_type_raw, sep = "_", into = c("cnv_type", "chr")) |>
  # remove extra chromosomes and cells not in this library
  dplyr::filter(
    !stringr::str_starts(chr, "chrMT"),
    !stringr::str_starts(chr, "chrGL"),
    library_id == params$library_id
  ) |>
  # set chr factor levels, with ones used for validation first
  dplyr::mutate(
    chr = stringr::str_remove(chr, "^chr"),
    chr = factor(chr, levels = 1:22),
    chr = forcats::fct_relevel(chr, as.character(validation_df$chromosome))
  ) |>
  # finally, join in the cell types
  dplyr::inner_join(
    has_cnv_df |>
      dplyr::select(barcodes, celltype_n)
  )

prop_cnv_df
```

```{r}
# separate gain and loss dfs
loss_df <- prop_cnv_df |>
  dplyr::filter(cnv_type == "loss")
gain_df <- prop_cnv_df |>
  dplyr::filter(cnv_type == "dupli")
```

We'll make density plots of the predicted CNV events across chromosomes, considering gain and loss separately.
The panels will be ordered first by the chromosomes with known CNVs, and then in order for the rest of the chromosomes.
Each line in the density plot represents a different cell type grouping.

```{r fig.height=12, fig.width=14}
ggplot(loss_df) +
  aes(x = proportion, color = celltype_n) +
  geom_density() +
  facet_wrap(
    vars(chr),
    scales = "free",
    ncol = 4 # this way the first row is the validation ones only
  ) +
  ggtitle("Proportion of CNV loss across chromosomes for each cell type group") +
  theme(strip.text = element_text(size = 8))
```

```{r fig.height=12, fig.width=14}
ggplot(gain_df) +
  aes(x = proportion, color = celltype_n) +
  geom_density() +
  facet_wrap(
    vars(chr),
    scales = "free",
    ncol = 4 # this way the first row is the validation ones only
  ) +
  ggtitle("Proportion of CNV gain across chromosomes for each cell type group") +
  theme(strip.text = element_text(size = 8))
```
